---
title: "Simulating detection of COVID-19 with wastewater MGS"
draft: false
format:
  html:
    code-fold: false 
execute:
  cache: false
knitr:
  opts_chunk: 
    message: true
editor: source 
author: Dan Rice
date: 2025-08-25
date-modified: today
---

# Objective

To get a very rough answer to the question:
If a wastewater MGS surveillance system had been operational in 2020,
when would it have detected SARS-CoV-2 reads?

# Setup

```{r}
#| label: setup
#| code-fold: true
#| message: false

library(dplyr)
library(tidyr)
library(lubridate)
library(purrr)
library(ggplot2)
library(slider)

set.seed(20250703)

theme_set(theme_minimal())

# Colors borrowed from https://github.com/JLSteenwyk/ggpubfigs
wong_eight <- c(
  "#E69F00",
  "#56B4E9",
  "#009E73",
  "#F0E442",
  "#0072B2",
  "#D55E00",
  "#CC79A7",
  "#000000"
)
options(
  ggplot2.discrete.colour = function()
    scale_colour_manual(values = wong_eight)
)
options(
  ggplot2.discrete.fill = function()
    scale_fill_manual(values = wong_eight)
)
```

# Model

## Epidemic

- A single deterministic incidence curve for one population
- Exponential growth for two months starting from a single case
- NOT an SIR model, but exponential growth of incidence is an intermediate-time
  approximate solution to SIR

```{r}
simulate_epidemic <- function(
  population_size,
  doubling_time, # in days
  day_zero # date when incidence = 1 case
){
  growth_rate <- log(2) / doubling_time
  max_day <- 60
  tibble(
    day = 0:max_day,
    date = day_zero + day,
    # Exponential growth from a single case
    incidence = exp(day * growth_rate) / population_size,
    cumulative_incidence = cumsum(incidence),
    # slide_dbl includes the current day plus .before days before it.
    weekly_incidence = slide_dbl(incidence, sum, .before = 6, .complete = FALSE),
  )
}
```

## Detection

- Model reads as negative binomial with an expectation set by the
  weekly incidence, $RA_i(1%)$, and the number of reads sequenced per sample.
- Detection is based on observing at least `detection_reads` cumulative reads across all samples.

```{r}
simulate_detection <- function(
    epi_sim, # output of `simulate_epidemic`
    num_sites,
    samples_per_site_per_day,
    ra_i_01, # expected relative abundance at 1% weekly incidence
    phi, # overdispersion parameter for read counts
    detection_reads, # number of cumulative reads required to detect virus
    delay_days, # days of delay between sample collection and detection
    reads_per_sample,
    ... # absorb unused parameters. allows us to use pmap with the full input data
) {
  mu <- ra_i_01 * 100
  num_reps <- 100
  epi_sim %>%
    # For each day in the simulation, simulate num_reps replicates
    # of the daily sampling process
    cross_join(
      expand_grid(
        rep = 1:num_reps,
        site = 1:num_sites,
        sample = 1:samples_per_site_per_day,
        )
    ) %>%
    # Simulate reads
    mutate(
      expected_reads = weekly_incidence * mu * reads_per_sample,
      reads = rnbinom(n(), size = phi, mu = expected_reads),
    ) %>%
    # Add sites and samples together
    summarise(
      reads = sum(reads),
      .by = day:rep,
    ) %>%
    # Get cumulative reads per day
    mutate(
      cumulative_reads = cumsum(reads),
      .by = rep,
    ) %>%
    # Simulate detection
    summarise(
      threshold_day = day[which.max(cumulative_reads >= detection_reads)],
      detection_day = threshold_day + delay_days,
      .by = rep,
    ) %>%
    left_join(
      epi_sim,
      by = join_by(detection_day == day),
    ) %>%
    # Return median detection date and cumulative incidence across reps 
    summarise(
      across(c(date, cumulative_incidence), median),
    )
}
```

# Parameters

## Epidemic parameters

- Start with the median estimated number of US infections from the SI of
  [this paper](https://www.nature.com/articles/s41586-021-04130-w).
- Fit an exponential growth curve to the estimates to get growth rate
  and projected day when the first person was infected.

```{r}
estimated_infections <- tribble(
 ~date, ~infections,
 "1/21/2020", 7,
 "1/22/2020", 8,
 "1/23/2020", 11,
 "1/24/2020", 13,
 "1/25/2020", 16,
 "1/26/2020", 18,
 "1/27/2020", 21,
 "1/28/2020", 26,
 "1/29/2020", 32,
 "1/30/2020", 39,
 "1/31/2020", 47,
 "2/1/2020", 55,
 "2/2/2020", 62,
 "2/3/2020", 75,
 "2/4/2020", 92,
 "2/5/2020", 111,
 "2/6/2020", 136,
 "2/7/2020", 165,
 "2/8/2020", 193,
 "2/9/2020", 217,
 "2/10/2020", 261,
 "2/11/2020", 322,
 "2/12/2020", 387,
 "2/13/2020", 476,
 "2/14/2020", 573,
 "2/15/2020", 664,
 "2/16/2020", 749,
 "2/17/2020", 887,
 "2/18/2020", 1072,
 "2/19/2020", 1301,
 "2/20/2020", 1566,
 "2/21/2020", 1888
) %>%
  mutate(
    date = mdy(date),
    day = as.numeric(date - min(date)),
    total_infections = cumsum(infections),
    )

model <- lm(log(infections) ~ day, data = estimated_infections)
doubling_time <- log(2) / coef(model)[[2]]
# log(1) = a + b t_0
# day_zero = first_date + t_0
day_zero <- min(estimated_infections$date) - coef(model)[[1]] / coef(model)[[2]]
print(doubling_time)
print(day_zero)
```

Note this is kind of a long doubling time compared to other estimates I've seen,
but I think it's a good idea to stick to this concrete estimate.

Calculate incidence curve:

```{r}
epi_params <- expand_grid(
    population_size = 330e6, # US population
    day_zero = day_zero,
    doubling_time = doubling_time,
  ) %>%
  mutate(epi_sim = pmap(., simulate_epidemic))
```

Check incidence curve against estimate:

```{r}
epi_params %>%
  unnest(epi_sim) %>%
  filter(date < ymd(20200301)) %>%
  mutate(total_infections = cumulative_incidence * 330e6) %>%
  ggplot(aes(date, total_infections)) +
  geom_line() +
  geom_point(data = estimated_infections)
```

Line is our curve; points are the input estimate.

## Sensitivity parameters

- Use the Rothman $RA_i(1\%)$ for SARS-CoV-2 as the baseline.
- Also include a 10x increase as a proxy for airplane waste or an improvement in tech.
- Require 5 reads for detection

```{r}
ra_i_01_rothman <- 6e-8 # posterior median of Rothman SARS-CoV-2 in Grimm et al 2023
sensitivity_params <- tribble(
  ~sensitivity, ~ra_i_01,
  "rothman", ra_i_01_rothman,
  "10x rothman", ra_i_01_rothman * 10,
) %>%
  mutate(
    phi = 1, # prior mode for wastewater phi in the swab p2ra blog post
    detection_reads = 5,
  )
```

## Cost and sequencing parameters

- Annual budget: $10M
- Non-sequencing costs from this NAO quote [private doc](https://docs.google.com/spreadsheets/d/1gdml2vNYR6tKPi34nrClhaPZ_4zpMeCED_0gY_tXLcs/edit?gid=1614122669#gid=1614122669)

```{r}
budget_params <- expand_grid(
  annual_budget = 10e6,
  budget_per_day = annual_budget / 365,
  cost_per_sample = 600, # ~$300 for consumables, ~$300 for wet lab labor
  nonseq_cost_per_run = 2300, # ~$2K for wet lab labor, $300 dry lab
)

non_sequencing_delay <- 4 # shipping, wet lab, computation
sequencing_params <- tribble(
  ~flow_cell, ~reads_per_run, ~seq_cost_per_run, ~delay_days,
  "10B", 10e9, 12e3, non_sequencing_delay + 1,
  "25B", 25e9, 16e3, non_sequencing_delay + 2, # Takes an extra day to run
)
```

## Sampling parameters

- Sample daily from a number of sites.
- We don't want to oversequence the libaries, so say that after 1B reads sequenced
  from a sample, we don't get any new information.
- To get around this, we can sequence multiple samples from the same location on the same day.
  How many samples should we sequence per location per day?
- Our budget also allows us to do at least one sequencing run per day. How many should we do?
- Solve the constrained optimization problem:
  - Maximize the number of useful reads we get while staying under-budget.
  - Vary the number of runs per day and the number of samples per location per day.

```{r}
max_reads_per_sample <- 1e9 # Can't get useful information from more than 1B reads per sample
sampling_params <- budget_params %>%
  cross_join(sequencing_params) %>%
  cross_join(
    expand_grid(
      num_sites = c(5, 10),
      samples_per_site_per_day = seq(1, 10),
      runs_per_day = seq(1, 2),
      )
    ) %>%
  mutate(
    cost_per_run = nonseq_cost_per_run + seq_cost_per_run,
    samples_per_day = num_sites * samples_per_site_per_day,
    samples_per_run = samples_per_day / runs_per_day,
    cost_per_day = runs_per_day * cost_per_run + samples_per_day * cost_per_sample,
    reads_per_sample = pmin(reads_per_run / samples_per_run, max_reads_per_sample),
    reads_per_day = reads_per_sample * samples_per_day,
  ) %>%
  # Only consider schemes that are under budget
  filter(cost_per_day <= budget_per_day) %>%
  # Take the scheme with the maximum total effective reads per day
  slice_max(
    reads_per_day,
    by = 1:num_sites,
  ) %>%
  # Break ties by lowest cost per day
  slice_min(
    cost_per_day,
    by = 1:num_sites,
  )

sampling_params %>%
  select(flow_cell, num_sites, runs_per_day, samples_per_site_per_day, reads_per_sample) %>%
  print
```

Looks like we should always do one sequencing run, that the optimal number of samples per site
per day varies with the flow cell and number of sites, and that we'll end up oversequencing.

# Simulation results

Run the simulation for every parameter combination.

```{r}
params <- epi_params %>%
  cross_join(sampling_params) %>%
  cross_join(sensitivity_params)

sim_results <- params %>%
  mutate(simulation = pmap(., simulate_detection)) %>%
  unnest(simulation)
```

Detection date for different scenarios:

```{r}
#| code-fold: true
sim_results %>%
  ggplot(
    aes(
      x = date,
      y = sensitivity,
      color = flow_cell,
      shape = as.factor(num_sites),
      ),
    ) +
  geom_point(position = position_dodge(width = 0.1)) 
```

The 10x difference in sensitivity makes a big difference (as you'd expect).
The flow cell and number of sites do not.
Of course, our model doesn't account for geographic variation in spread,
so this is not surprising.

Detection cumulative incidence for different scenarios:

```{r}
#| code-fold: true
sim_results %>%
  ggplot(
    aes(
      x = cumulative_incidence,
      y = sensitivity,
      color = flow_cell,
      shape = as.factor(num_sites),
      ),
    ) +
  geom_point(position = position_dodge(width = 0.1)) +
  scale_x_log10() 
```

As a table:

```{r}
sim_results %>%
  select(sensitivity, num_sites, flow_cell, date, cumulative_incidence) %>%
  print
```